Foundations of ecological and evolutionary change

### 2025-08-04 foundations_of_ecoevo.Rmd
Line 99 reviewer suggested edit. The w was just missing here, so this was a typo

> **Box 1:** The Price equation is an abstract formula to represent evolutionary change. Formulated originally in the early 1970s by George Price [@Price1970; @Price1972], it postulates some basic properties that all evolutionary systems must satisfy: change over time, ancestor and descendant relations, and a character or phenotype [@Rice2004]. Using simple algebraic language, the Price equation represents evolutionary change with the predominant notation, $$\bar{w}\Delta\bar{z} = \mathrm{Cov}\left(w, z\right) + \mathrm{E}\left(w\Delta z\right).$$ In the above equation, $\Delta\bar{z}$ is the change in the average character value $z$ over a time step of arbitrary length, $w$ is an individual's fitness, and $\bar{w}$ average population fitness. On the right-hand side of the equation, the first term is the covariance between a character value $z$ and fitness $w$, which reflects $\bar{z}$ change attributable to differential survival and reproduction. The second term is the expected value of $\Delta z$, which reflects the extent to which offspring deviate from parents in $z$ [@Rice2004; @Okasha2006; @Frank2012]. A more specific version of the covariance term was already known within the quantitative and population genetics tradition [@Robertson1966], usually representing the action of natural selection. The Price equation adds an expectation term and abstracts away from any specific mechanisms of replication or reproduction, or mechanisms of inheritance. Its definitional nature and lack of substantive biological assumptions has been portrayed both as a strength [@Rice2004; @Frank2012; @Luque2016; @Baravalle2022], and its greatest weakness. The abstract nature of the Price equation places it at the top of the hierarchy of fundamental theorems of evolution from which the rest (Robertson's theorem, Fisher's fundamental theorem, breeder's equation, Hamilton's rule, adaptive dynamics, etc.) can be derived [@Queller2017; @Lehtonen2016; @Lehtonen2018; @Lehtonen2020a]. This abstractness is also key to developing a more general view of evolution [@Rice2020; @Luque2021; @Edelaar2023]. In contrast, some researchers consider the Price equation just a triviality (even tautological), and useless without further modelling assumptions [@VanVeelen2005; @VanVeelen2012]. The debate remains open [@VanVeelen2020; @Baravalle2024].

@@

Previous theory has produced mathematical synthesis between ecological and evolutionary equations [@Page2002; @Collins2009], but not established a unified conceptual framework for joint ecological and evolutionary change.
For example, @Page2002 demonstrate a mathematical equivalence between the Price equation and a general equation for community dynamics. 
In both equations, the relative frequencies of different ecological (species) or evolutionary (phenotype) variants map to their respective ecological (population growth) or evolutionary (fitness) change in a way that is mathematically equivalent.
@Collins2009 use the Price equation to build a general quantitative framework that partitions phenotypic change into ecological, evolutionary, and physiological components.
The average change in a community-wide trait (e.g., carbon uptake) is exactly and generally partitioned into change attributable to changes in species composition, expected evolution of lineages within species, and expected change in lineage physiology across all species.
Both @Page2002 and @Collins2009 use relative species abundances, meaning that species frequencies must sum to one.
This is useful for demonstrating mathematical equivalence [@Page2002] or expected change in a community-wide trait [@Collins2009], but a different approach is needed to recover absolute change in population size and thereby recover a joint foundation for ecology and evolution.

>>>>

> **Box 1:** The Price equation is an abstract formula to represent evolutionary change. Formulated originally in the early 1970s by George Price [@Price1970; @Price1972], it postulates some basic properties that all evolutionary systems must satisfy: change over time, ancestor and descendant relations, and a character or phenotype [@Rice2004]. Using simple algebraic language, the Price equation represents evolutionary change with the predominant notation, $$\bar{w}\Delta\bar{z} = \mathrm{Cov}\left(w, z\right) + \mathrm{E}\left(w\Delta z\right).$$ In the above equation, $\Delta\bar{z}$ is the change in the average character value $z$ over a time step of arbitrary length, $w$ is an individual's fitness, and $\bar{w}$ average population fitness. On the right-hand side of the equation, the first term is the covariance between a character value $z$ and fitness $w$, which reflects $\bar{z}$ change attributable to differential survival and reproduction. The second term is the expected value of $w \Delta z$, which reflects the extent to which offspring deviate from parents in $z$ [@Rice2004; @Okasha2006; @Frank2012]. A more specific version of the covariance term was already known within the quantitative and population genetics tradition [@Robertson1966], usually representing the action of natural selection. The Price equation adds an expectation term and abstracts away from any specific mechanisms of replication or reproduction, or mechanisms of inheritance. Its definitional nature and lack of substantive biological assumptions has been portrayed both as a strength [@Rice2004; @Frank2012; @Luque2016; @Baravalle2022], and its greatest weakness. The abstract nature of the Price equation places it at the top of the hierarchy of fundamental theorems of evolution from which the rest (Robertson's theorem, Fisher's fundamental theorem, breeder's equation, Hamilton's rule, adaptive dynamics, etc.) can be derived [@Queller2017; @Lehtonen2016; @Lehtonen2018; @Lehtonen2020a]. This abstractness is also key to developing a more general view of evolution [@Rice2020; @Luque2021; @Edelaar2023]. In contrast, some researchers consider the Price equation just a triviality (even tautological), and useless without further modelling assumptions [@VanVeelen2005; @VanVeelen2012]. The debate remains open [@VanVeelen2020; @Baravalle2024].

@@
